---
title: "Assignment 2 - part 1"
author: "Pablo"
---

# 0. Packages

```{r}
library(sf) # simple features' library
library(spData) # library of spatial datasets
library(tidyverse) # dplyr, ggplot, ...
library(here)  # for setting relative paths
```


# 1. Combination of the `world` with more data

Combination with:

- Population (point) data (do not use rasters!)
- Ports, airports, etc.

## 1.1. World dataset

Loading the world dataset:

```{r}
gmsf_world <- world
head(gmsf_world)
```

We keep only the variables that interest us for this task:

```{r}
gmsf_world_sel <- gmsf_world %>% select(iso_a2, name_long, continent, type, pop, geom)
View(gmsf_world_sel)
rm(gmsf_world)
```

*Note*: just using the world dataset to produce a map of total population by country is not enough, as there are countries with missing values (see below).

```{r}
# Plotting it with ggplot:
ggplot(gmsf_world_sel) +
  geom_sf(aes(fill=pop))
```

## 1.2. Population data by location

Source of the population data: https://www.naturalearthdata.com/downloads/10m-cultural-vectors/10m-populated-places/

```{r}
path_population_shp <- here("geo_assig2/map_1/populated_places", "ne_10m_populated_places.shp")
gmsf_population <- st_read(path_population_shp)
head(gmsf_population)
colnames(gmsf_population)
```

Again, we only select those variables that may interest us:

```{r}
gmsf_population_sel <- gmsf_population %>% 
  select(FEATURECLA, NAME, NAMEASCII, SOV0NAME, ADM0NAME, ISO_A2, POP_MAX, POP_MIN, geometry)
head(gmsf_population_sel)
rm(gmsf_population)
```

## 1.3. Population data by country

Source of the population by country data: https://www.naturalearthdata.com/downloads/10m-cultural-vectors/10m-admin-0-countries/

```{r}
path_population_countries_shp <- here("geo_assig2/map_1/population_by_countries", "ne_10m_admin_0_countries.shp")
gmsf_population_by_countries <- st_read(path_population_countries_shp)
head(gmsf_population_by_countries)
colnames(gmsf_population_by_countries)
```

```{r}
gmsf_population_by_countries_sel <- gmsf_population_by_countries %>%
  select(SOVEREIGNT, ADMIN, ISO_A2, CONTINENT, POP_EST, POP_YEAR, geometry)
head(gmsf_population_by_countries_sel)
rm(gmsf_population_by_countries)
```

```{r}
rows_with_na <- gmsf_population_by_countries_sel %>% filter(is.na(POP_EST))
head(rows_with_na)
```

There are no missing values for the population field (an improvement with respect to the `world` dataset).

## 1.4. Ports data

Source of the ports data: https://www.naturalearthdata.com/downloads/10m-cultural-vectors/ports/

```{r}
path_ports_shp <- here("geo_assig2/map_1/ports", "ne_10m_ports.shp")
gmsf_ports <- st_read(path_ports_shp)
head(gmsf_ports)
colnames(gmsf_ports)
```

Select only the values that may be useful:

```{r}
gmsf_ports_sel <- gmsf_ports %>% select(featurecla, name, ne_id, geometry)
head(gmsf_ports_sel)
rm(gmsf_ports)
```

## 1.5. Airports data

Source of the airports data: https://www.naturalearthdata.com/downloads/10m-cultural-vectors/airports/

```{r}
path_airports_shp <- here("geo_assig2/map_1/airports", "ne_10m_airports.shp")
gmsf_airports <- st_read(path_airports_shp)
head(gmsf_airports)
colnames(gmsf_airports)
```

Select only the features that may be useful:

```{r}
gmsf_airports_sel <- gmsf_airports %>% select(featurecla, type, name, abbrev, ne_id, geometry)
head(gmsf_airports_sel)
rm(gmsf_airports)
```

# 2. Maps

## 2.1. Map of total population by country

```{r}
ggplot(gmsf_population_by_countries_sel) +
  geom_sf(aes(fill=POP_EST)) +  # Fill countries by population estimates
  scale_fill_distiller(palette = "Spectral",  # Setting color legend
                       name = "Population Estimate") +
  labs(title = "World Population by Country",
       subtitle = "Population estimates across countries",
       caption = "Data Source: Natural Earth (2025)",
       )
```

Note, however, that the population estimates correspond to different years, so the current situation may differ slightly.

## 2.2. Histogram of country population distribution by continent

```{r}
ggplot(gmsf_population_by_countries_sel, aes(POP_EST)) +
  geom_histogram(bins = 50, fill = "purple", color = "white") +
  facet_wrap(~ CONTINENT, scales = "free_x") + # We use facet_wrap to create a figure with several histograms grouped by continent
  labs(title = "Population Histogram by Continent (different population scale)",
       x = "Population Estimate",
       y = "Frequency")
```

```{r}
ggplot(gmsf_population_by_countries_sel, aes(POP_EST)) +
  geom_histogram(bins = 50, fill = "purple", color = "white") +
  facet_wrap(~ CONTINENT, scales = "fixed") + # We use facet_wrap to create a figure with several histograms grouped by continent
  labs(title = "Population Histogram by Continent (same population scale)",
       x = "Population Estimate",
       y = "Frequency")
```

## 2.3. Histogram of (country-level) average distances between locations and ports or airports by continent

For creating these histograms, we need to:
1. First, compute the distances of each location to all the airports and ports within the same country (if computationally feasible, if not limit to top 20 locations by country).
2. Second, keep just the shortest distance of each location within a country to an airport and a port.
3. Third, compute the average distances (country-level) of each location to an airport and a port.
4. Fourth, plot the histogram with the average distances by continent.

### Preliminary cleaning

First, we filter out those airports which only have a military or spaceport purpose:

```{r}
unique(gmsf_airports_sel$type)
```

We remove those airports which just have a military or spaceport purpose, since they won't be an indicator for how well connected a location is.

```{r}
gmsf_airports_slice <- gmsf_airports_sel %>%
  filter(!(type %in% c("military", "military mid", "military major", "spaceport")))
unique(gmsf_airports_slice$type)
```

### Step 1: Compute distances of (populated) locations to infrastructures in the same country

What is the number of calculations that we would have to do if we wanted to find the distances of all locations from all the infrastructures?

```{r}
print(paste("Number of calculations for airports:", dim(gmsf_airports_slice)[1] * dim(gmsf_population_sel)[1]))
print(paste("Number of calculations for ports:", dim(gmsf_ports_sel)[1] * dim(gmsf_population_sel)[1]))
```

So, in the case of airports, we will have to do 6,416,908 distance calculations. For ports, 7,936,702.

Calculations for ports:

```{r}
ports_distance <- st_distance(gmsf_population_sel, gmsf_ports_sel, by_element = F)
```

The resulting matrix has `dim(gmsf_population_sel)[1]` rows and `dim(gmsf_ports_sel)[1]` columns. 

Each element [i, j] in the matrix represents the distance between the i-th geometry in the first object and the j-th geometry in the second object.

Calculations for airports:

```{r}
airports_distance <- st_distance(gmsf_population_sel, gmsf_airports_slice, by_element = F)
```

### Step 2: Keep the shortest distance of each location with each type of infrastructure

Since the resulting matrices have `dim(gmsf_population_sel)[1]` rows and `dim(gmsf_ports_sel)[1]` columns, we can select the minimum distance of each location with each infrastructure by selecting the minimum value of each row. 

```{r}
dim(airports_distance)
dim(ports_distance)
```

Now, we compute the minimum distance for each row (location) with each type of infrastructure, and save it as a vector:

```{r}
min_airports_distances <- apply(
  X = airports_distance, MARGIN = 1, FUN = min, na.rm = TRUE
)
min_ports_distances <- apply(
  X = ports_distance, MARGIN = 1, FUN = min, na.rm = TRUE
)
```

We convert the vectors to km for making the units more readable:

```{r}
min_airports_distances_km <- min_airports_distances / 1000
min_ports_distances_km <- min_ports_distances / 1000
```

```{r}
head(min_airports_distances_km)
```

Below, we bind the minimum distance vectors of each type of infrastructure as new columns of the `gmsf_population_sel` `sf`.

Since the order of the rows in `gmsf_population_sel` remains unchanged during the distance matrix calculation, we can directly bind the minimum distance vector as a new column.

```{r}
gmsf_population_sel$min_distance_airport_km <- min_airports_distances_km
gmsf_population_sel$min_distance_port_km <- min_ports_distances_km
head(gmsf_population_sel)
```

### Step 3: Compute average distances (country-level) of each location to each type of infrastructure

Now, the idea is to group by the sovereign country name and compute the average distance of each location with each type of infrastructure, ports and airports:

```{r}
# Group by country (using the ISO_A2 country code)
avg_country_dist_airport <- gmsf_population_sel %>%
  group_by(ISO_A2) %>%
  summarise(avg_dist_airport_km = mean(min_distance_airport_km, na.rm = T))

avg_country_dist_port <- gmsf_population_sel %>%
  group_by(ISO_A2) %>%
  summarise(avg_dist_port_km = mean(min_distance_port_km, na.rm = T))
```

```{r}
head(avg_country_dist_airport)
head(avg_country_dist_port)
```

Now, we drop the geometries from the data frames containing the average distances and merge the results with the `gmsf_population_by_countries_sel` data frame, which has the same names for the sovereign countries and also contains the country polygons:

```{r}
# First, we drop the geometries
avg_country_dist_airport <- st_drop_geometry(avg_country_dist_airport)
avg_country_dist_port <- st_drop_geometry(avg_country_dist_port)

# Second, we do a left join on the distance data frames, in order to
# have the polygons of the world data frame
gmsf_avg_distances <- left_join(
  x = gmsf_population_by_countries_sel,
  y = avg_country_dist_airport,
  by = "ISO_A2")
gmsf_avg_distances <- left_join(
  x = gmsf_avg_distances,
  y = avg_country_dist_port,
  by = "ISO_A2")
```

```{r}
head(gmsf_avg_distances)
```

### Step 4: Plotting histograms by type of infrastructure

Note that not all locations in the world are included in the dataset that has been used, so the histograms may not accurately represent reality.

```{r}
ggplot(gmsf_avg_distances, aes(avg_dist_airport_km)) +
  geom_histogram(bins = 50, fill = "red", color = "white", na.rm = T) +
  facet_wrap(~ CONTINENT, scales = "fixed") + # We use facet_wrap to create a figure with several histograms grouped by continent
  labs(title = "Histogram of Country-level Average Distances to Closest Airport by Continent",
       x = "Average Distance (km)",
       y = "Frequency")
```

```{r}
ggplot(gmsf_avg_distances, aes(avg_dist_port_km)) +
  geom_histogram(bins = 50, fill = "blue", color = "white", na.rm = T) +
  facet_wrap(~ CONTINENT, scales = "fixed") + # We use facet_wrap to create a figure with several histograms grouped by continent
  labs(title = "Histogram of Country-level Average Distances to Closest Port by Continent",
       x = "Average Distance (km)",
       y = "Frequency")
```

### Additional: maps of country-level average distances to the closest ports and airports

```{r}
ggplot(gmsf_avg_distances) +
  geom_sf(aes(fill=avg_dist_airport_km)) +  # Fill countries by avg distances
  scale_fill_distiller(palette = "Spectral",  # Setting color legend
                       name = "Average Distance to Airport") +
  labs(title = "Average Distance to Closest Airport",
       subtitle = "Country-level average distances to the closest airport, in km (using the arithmetic mean for populated locations)",
       caption = "Data Source: Natural Earth (2025)",
       )
```

```{r}
ggplot(gmsf_avg_distances) +
  geom_sf(aes(fill=avg_dist_port_km)) +  # Fill countries by avg distances
  scale_fill_distiller(palette = "Spectral",  # Setting color legend
                       name = "Average Distance to Port") +
  labs(title = "Average Distance to Closest Port",
       subtitle = "Country-level average distances to the closest port, in km (using the arithmetic mean for populated locations)",
       caption = "Data Source: Natural Earth (2025)",
       )
```